---
title: "Analysis of the Harmful Effects of Storms to the Health and Economy of the US Population"
author: "Md. Mizanur Rhaman"
date: "9/12/2020"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1: Synopsis
This is a report of exploring the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database and analysis corresponding to most disaster storm event impacts in United States on public health and economic consequences. The database tracks the characteristics of major storms and weather events in the United States. The data was collected during the period from 1950 through 2011. In specific there are two questions that have been tried to answer through this analysis:

Across the United States, which type of events are most harmful with respect to population health?
Across the United States, which type of events have the greatest economic consequences?
After the analysis of storm data events it was found that Tornado causes the most public health impact, while Flood causes the most economic impact in relation to crop and property damage.
  
  
## 2: Loading and Processing the Raw Data
The [Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp) on which the analysis is performed, provided by [National Climatic Data Center](http://www.ncdc.noaa.gov/). The compressed data can be downloaded from the [source URL](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). Furthermore the documentation details of the data can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

### 2.1: Downloading the raw data file into the current folder
```{r downloading, cache=TRUE, echo=TRUE}
file_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(file_url, destfile = "storm_data.csv.bz2", method = "curl")
```

### 2.2: Extracting the data into a dataframe. Then converting it into a data.table
```{r extraction, echo=TRUE}
library(data.table)
library(dplyr)
library(ggplot2)

dataset <- read.csv("storm_data.csv.bz2")
# Converting data.frame to data.table
dataset <- as.data.table(dataset)

```

### 2.3: Examining the dataset and preparing it for analysis
```{r examining and subsetting, echo=TRUE}

# Reading all the variable names in the dataset
names(dataset)

# Subsetting the variables we will need for our analysis:
dataset <- subset(x = dataset,
                  subset = (EVTYPE != "?" & 
                            (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)),
                  select = c(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```



### 2.4: Converting the data values into their numeric forms
```{r numeric conversion, echo=TRUE}

# Converting the PROPDMGEXP and CROPDMGEXP values into numeric values for calculations:
# Inspecting the levels of PROPDMGEXP and CROPDMGEXP
levels(dataset$PROPDMGEXP)
levels(dataset$CROPDMGEXP)

# Create a lookup table with the numeric values corresponding to each level of PROPDMGEXP and CROPDMGEXP
prop_lookup <- data.table(PROPDMGEXP = levels(dataset$PROPDMGEXP), 
                          PROPDMGEXPNum = c(10^0, 10^0, 10^0, 10^0, 10^0, 10^1, 10^2, 
                                            10^3, 10^4, 10^5, 10^6, 10^7, 10^8, 10^9, 
                                            10^2, 10^2, 10^3, 10^6, 10^6))

crop_lookup <- data.table(CROPDMGEXP = levels(dataset$CROPDMGEXP), 
                          CROPDMGEXPNum = c(10^0, 10^0, 10^0, 10^2, 10^9, 10^3, 10^3, 
                                            10^6, 10^6) )

# Merging the two lookup values into the dataset
dataset <- merge.data.table(dataset, prop_lookup, by = "PROPDMGEXP")
dataset <- merge.data.table(dataset, crop_lookup, by = "CROPDMGEXP")
remove(crop_lookup, prop_lookup)


# Calculating the Property Damage Cost and Crop Damage Cost in their respective numeric values:
dataset <- mutate(dataset, PROPDMGCost = PROPDMG * PROPDMGEXPNum, CROPDMGCost = CROPDMG * CROPDMGEXPNum)

# Dropping the unnecessary columns from the dataset
dataset <- select(dataset, -PROPDMG, -PROPDMGEXP, -PROPDMGEXPNum, -CROPDMG, -CROPDMGEXP, -CROPDMGEXPNum)
dataset <- as.data.table(dataset)
```



### 2.5: Calculating the total damage, fatalities and injuries happened
```{r calculation of totals, echo=TRUE}
total_damage <- dataset[, .(PROPDMGCost = sum(PROPDMGCost), CROPDMGCost = sum(CROPDMGCost), Totals = sum(PROPDMGCost) + sum(CROPDMGCost)), by = .(EVTYPE)]
total_damage <- total_damage[order(-Totals)]
total_damage <- total_damage[1:10, ]
head(total_damage, 10)


# Calcuating Total Fatalities and Injuries
total_injuries <- dataset[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), Totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]
total_injuries <- total_injuries[order(-Totals)]
total_injuries <- total_injuries[1:10, ]
head(total_injuries, 10)
```


## 3: Results


### 3.1: Events that are Most Harmful to Population Health
```{r health results, echo=TRUE}
# Melting the total_injuries data.table to put it easily in the bar graph
Harmful_Events <- melt(total_injuries, id.vars="EVTYPE", variable.name = "Type_of_Event")

# Creating the barplot
ggplot(Harmful_Events, aes(x=reorder(EVTYPE, -value), y=value)) + 
    geom_bar(stat="identity", aes(fill=Type_of_Event), position="dodge") + 
    ylab("Frequency Count") +
    xlab("Event Type") + 
    theme(axis.text.x = element_text(angle=45, hjust=1)) +
    ggtitle("Top 10 Harmful Events for US Population") + 
    theme(plot.title = element_text(hjust = 0.5))
```  
  
> According to the analysis, **Tornado** is the natural disaster which causes most harm to the US population. It is the most fatalities and injuries causing event.



### 3.2: Events that have the Greatest Economic Consequences
```{r EconConsequences}
# Melting the total_damage data.table to put it easily in the bar graph 
Economic_Consequences <- melt(total_damage, id.vars="EVTYPE", variable.name = "Type_of_Damage")

# Creating the barplot
ggplot(Economic_Consequences, aes(x=reorder(EVTYPE, -value), y=value)) +
    geom_bar(stat="identity", aes(fill=Type_of_Damage), position="dodge") +
    ylab("Cost (in dollars)") +
    xlab("Event Type") + 
    theme(axis.text.x = element_text(angle=45, hjust=1)) +
    ggtitle("Top 10 Events causing Economic Disasters in US") +
    theme(plot.title = element_text(hjust = 0.5))
```  
  
> From the graph, we can say **Flood** is the most harmful event which causes the most economic disaster in the US. Here, **Property damages** are higher than **Crop damages**.

